---
title: "MR-Clust vignette"
output: 
    html_document:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{mrclust}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width=5, 
  fig.height= 4
)
```

```{r setup}
library(mrclust)
```


# Introduction


MR-Clust is used to investigate the presence of clustered heterogeneity in
Mendelian randomization analyses. Clustered heterogeneity is identified by the
presence of two or more distinct clusters of genetic variants - such that
genetic variants within a cluster recover similar estimates of the causal effect
of the risk-factor on the outcome, i.e. their ratio-estimates are similar in
direction, magnitude and precision. A cluster will typically comprise of two or
more genetic variants and the ratio-estimates and standard errors of these
variants are then used to compute the cluster centres. A cluster might represent
a distinct pathway with which the risk-factor is related to the outcome and
hence clustered heterogeneity is interesting to investigate as the identity
of the genetic variants in the clusters may reveal information about the risk
factor and how it relates to the outcome. 

MR-Clust performs likelihood based clustering on a sample of ratio-estimates and
associated standard error information. The MR-Clust mixture-model automatically
accounts for the possibility that some or all of the ratio-estimates may be
drawn under two distributions - the results from which we do not wish to
interpret - (i) a null distribution, in which the ratio-estimates are centred
around zero and; (ii) a junk distribution, which aims to model highly dispersed
ratio-estimates which are considered outliers in the sample. In some situations,
the junk-component may fail to capture all outliers in the sample, instead these
observations may appear in clusters comprising a single gnetic variant. Unless
there is some underlying motivation for doing so, we do not recommend
interpreting results from clusters containing one (or a small number of)
genetic variants.


# Installation


```{r, echo=F}
options(warn=-1)
```

```{r, eval=F}
install.packages("devtools", repos='http://cran.us.r-project.org')
library(devtools)
install_github("cnfoley/mrclust", build_vignettes = TRUE)
```

# Using MR-Clust 

We introduce the MR-Clust package by allowing users to replicate an analysis
from our manuscript. 

## Replicating the SBP-CAD applied analysis

Here we replicate the results from the MR-Clust manuscript: assessing the
presence of clustered heterogeneity in the analysis of systolic blood pressure
(SBP) and it's effects on coronary artery disease (CAD) risk. We begin by
loading the package and the SBP/CAD summary data. These data are avaiable with
the package and can be found by typing

```{r}
library(mrclust)
sbp_cad <- mrclust::SBP_CAD
head(sbp_cad)
```

Data from the other BP trait analyses, i.e. the diastolic blood pressure (DBP) and pulse pressure (PP) analyses, can be accessed similarly:

```{r}
# dbp_cad <- mrclust::DBP_CAD
# pp_cad <- mrclust::PP_CAD
```

In the combined systolic blood pressure (SBP) and coronary artery disease (CAD) dataset there are 8 variables: chromosome position, "chr.pos"; RSID; genetic associations with risk-factor SBP, "bx"; standard error of the genetic association with risk-factor SBP, "bxse"; genetic associations with outcome CAD, "by"; standard error of the genetic association with outcome CAD, "byse" and; the A1 and A2 alleles, "a1" and "a2". A description of these data is found by typing:

```{r}
?SBP_CAD
```

The function "mrclust" (minimally) requires the following summary data information: 

```{r}
##############################################################
# estimates of the G-X associations and their standard errors
##############################################################

bx = sbp_cad$bx;
bxse = sbp_cad$bxse;

##############################################################
# estimates of the G-Y associations and their standard errors
##############################################################

by = sbp_cad$by;
byse = sbp_cad$byse;

###########################################
# ratio-estimates and their standard errors
###########################################

ratio_est = by/bx;
ratio_est_se = byse/abs(bx);

#####################
# Names for the snps
#####################
snp_names = sbp_cad$chr.pos;
```

## Running an MR-Clust analysis 


To run a default MR-Clust analysis we use the "mr_clust_em" function (which employs algorithm 1 in the MR-Clust manuscript):      

```{r}
############################
########### SBP-CAD analysis
############################

res_em = mr_clust_em(theta = ratio_est, theta_se = ratio_est_se, bx = bx,
                     by = by, bxse = bxse, byse = byse, obs_names = snp_names);

```

The function returns a list whose elements summarise the output of the MR-Clust algorithm in 5 ways, which we discuss in turn. The names of these 5 summaries are found via:

```{r}
names(res_em);
```


## Allocation probabilities and the best clustering of the data

The "results" variable is a list containing two data tables "results\$all" and "results\$best". Both tables provide summaries of the ratio-estimates and standard errors for each genetic variant. In addition to this, data table "all" contains information on the allocation probability of each variant being assigned to one of the K+2 possible clusters; data table "best", however, presents information on the largest allocation probability for each variant only, i.e. each variant is assigned to a single (most probable) cluster. Hence, the results in data table "best" are considered the best clustering of the data:


```{r}
#############################
# results contain two tables:
#############################

names(res_em$results);

####################################################################################################
# first few rows of table containing summaries for the genetic variants allocated to "all" clusters 
####################################################################################################

head(res_em$results$all);

#######################################################################################################
# first few rows of table containing summaries for the genetic variants allocated to the "best" cluster
#######################################################################################################

head(res_em$results$best);

```


The results reveal that MR-Clust identified K=6 clusters in the SBP-CAD data: a null cluster, junk cluster and four additional clusters:


```{r}
######################
# Identified clusters
######################

clusters = unique(res_em$results$best$cluster_class); # same answer if we type unique(res_em$results$all$cluster_class)
clusters[order(clusters)];

######################
# number of clusters K
######################

K = length(clusters);
K;
```

### Summarising variants by allocation probability within each cluster


MR-Clust allows users to stratify the variants in a cluster by deciles of the
allocation probability, which might help when aiming to prioritise the variants
with high quality allocation probabilities for follow-up analyses. We do this 
using the second summary returned by MR-Clust: "cluster.membership". This is a
list - of length the number of clusters - which, for each cluster, batches the
variants, we use cluster 1 to illustrate the idea:

```{r}
######################
#  clusters
######################

names(res_em$cluster.membership)

###########################################################################
# Variants assinged to the first few deciles of the allocation probability 
###########################################################################

head(res_em$cluster.membership$cluster_1);
```
There are 5 variants assigned to cluster 1 with an allocation probability in the
range (0.9,1), there is 1 variant with an allocation prbability in the range
(0.8, 0.9) and no variants with allocation probabilities in the range (0.7,0.8).
We can single out the variants with the largest allocation probabilities by
typing

```{r}
#########################################################
# Variants assinged to cluster one with probability >0.9 
#########################################################

res_em$cluster.membership$cluster_1$pr_range_1_to_0.9;
```

## Plotting results


A scatter plot of results from the two-stage regression is returned by default
with each variant colour coded according to their cluster assignment (using the
best clustering). This is found by typing

```{r}
###################################
# Scatter plot of the fitted model  
###################################

plot.sbp.best = res_em$plots$two.stage + ggplot2::xlim(0, max(abs(bx)+2*bxse)) + ggplot2::xlab("Genetic association with SBP") + ggplot2::ylab("Genetic association with CAD") + ggplot2::ggtitle("");

##############
# plot output
##############

plot.sbp.best
```

As noted from our manuscript, the quality of the best clustering of the data typically improves when considering variants assigned to a cluster with a large allocation probability and for clusters containing several variants. For ease, we will assume that a large allocation probability is above 0.8 and several variants means 4 or more. We can make use of the MR-Clust function "pr.clust", to remove varinats which do not meet these criteria:      

```{r}
#################################################################################################
# Variants allocated to clusters with: (i) at least 4 variants and; (ii) allocation probability >=0.8 
#################################################################################################

res80 = mrclust::pr.clust(dta = res_em$results$best, prob = 0.8, min.obs =  4);
```


We can then re-plot the data using the results above and using the MR-Clust (two-stage regression estimate) plotting function "two.stage.plot", e.g. 


```{r}
##############################################################
# Update summary effects using the prioritised set of variants  
##############################################################

      keep80 = which(snp.names %in% res80$observation);
      bx80   = bx[keep80];
      bxse80 = bxse[keep80];
      by80   = by[keep80];
      byse80 = byse[keep80];
      snp.names80 = snp.names[keep80];

plot.sbp.pr80 = two.stage.plot(res = res80, bx = bx80, by = by80, bxse = bxse80, byse = byse80, obs.names = snp.names80) + ggplot2::xlim(0, max(abs(bx80)+2*bxse80)) + ggplot2::xlab("Genetic association with SBP") + ggplot2::ylab("Genetic association with CAD") + ggplot2::ggtitle("");
# plot result
plot.sbp.pr80;
```


## Searching for traits associated with clusters of variants


MR-Clust allows users to search for traits associated with each of the variants in the non-null or junk clusters, to identify candidate mechanisms which might explain the presence of clustered heterogeneity in the data. To do this we make use of the variable "trait.search" in "rm_clust_em". Setting "trait.search=TRUE" will perform a trait search for each of the variants that are not in either the null or junk clusters,  "trait.search" calls the "phenoscanner" function in the "MendelianRandomization" R package (https://cran.r-project.org/web/packages/MendelianRandomization/MendelianRandomization.pdf). This function takes in several variables which we use in MR-Clust also, i.e. the variables/parameters "catalogue" (default "GWAS"); proxies (deafult "None"); pvalue (coded "trait.pvalue"; default $10^{-5}$); build (genome build options 37 or 38, default 37); r2 (coded "proxy.r2", r2 = 0.8). See Phenoscanner (http://www.phenoscanner.medschl.cam.ac.uk/information/) for more information on the specification of these paramaters. 

```{r}
###################################################
# Re-run analysis with trait search option == TRUE
###################################################

res_em = mr_clust_em(theta=ratio.est, theta.se=ratio.est.se, bx=bx, by=by, bxse=bxse, byse=byse,
                       obs.names = snp.names, max.iter = 5e3, tol = 1e-5,
                       junk.mean = 0,
                       results.list = list("all", "best"),
                       cluster.membership = list(by_prob = 0.25, bound = 0.75),
                       plot.results = list("best", min.pr = 0.5),
                       trait.search = TRUE, trait.pvalue = 1e-5, proxy.r2 = 0.8,
                       catalogue = "GWAS", proxies = "None", build = 37);

```

In addition to the output decribed previously, "mr_clust_em" returns a variable "trait.search". This is accessed by typing "res_em$trait.search":

```{r}
##############################
# Access trait search results
##############################

res.search = res_em$trait.search;

####################
# variables returned
####################

names(res.search);
```

The variable "trait" contains information on the traits identified as having an association with at least one variant in "cluster" below the p-value "p.cut.off". The variable "associated.snps" denotes the total number of snps in the cluster which are associated with the trait and "total.snps" denotes the total number of variants in the "cluster" analysed. Note that "total.snps" may be smaller than the number of variants assinged to "cluster". This is beacuse MR-Clust performs a variant-trait search for variants assinged to a cluster in the upper most strata of "cluster.membership" (under the assumption that this will lead to better true positive trait associations). In our analysis above we specified that our "cluster.membership" variable be organised into allocation probability steps of size 0.25 from 1 to bound=0.75, i.e.

```{r}
############################################################
# Setting
# "cluster.membership = list(by_prob = 0.25, bound = 0.75)"
# in the argument of "mr_clust_em"
# returns:
############################################################

head(res_em$cluster.membership);
```

Only these variants (in the top strata and not from either the null or junk clusters) are considered in the trait search.

We continue by inspecting results from the largest cluster of traits in the data, which we find via the following: 

```{r}
###############################################
# Count the number of variants in each cluster 
###############################################

tab.clust = table(res_em$results$best$cluster_class)
tab.clust;

##########################################################
# Identify the cluster with the largest number of variants 
##########################################################

mx.clust = which.max(tab.clust[!names(tab.clust) %in% c("Null", "Junk")]);
clst.trt.srch = paste0("cluster_",mx.clust) 

```

For smaller values of p-value cutt-off, the number of traits identified can be large and many traits will be associated with a small number of variants in a cluster. To avoid inspecting these types of results, we can choose to inspect only results in which the difference between the total number of snps assessed "total.snps" and those identified as being associated with the trait "associated.snps" is `small', i.e. "total.snps"-"associated.snps" is small. For illustrative purposes, we take this to be less than half "total.snps", i.e.    

```{r}
##############################
# Half the total number of snp
##############################

diff.snps.tol = floor(tab.clust[mx.clust]/2);

##############################################
# Difference between total and associated snps
##############################################

diff.snps = as.numeric(res_em$trait.search$total.snps)-as.numeric(res_em$trait.search$associated.snps);

```

Finally some results!

```{r}
head(res_em$trait.search[which(diff.snps<=diff.snps.tol & res_em$trait.search$cluster==clst.trt.srch),]);
```

...unsurprisingly some BP traits have been identified. 
